Powering Hidden Markov Model by Neural Network based Generative Models

### 2020-02-04 manuscript/main.tex
fix typos

With such a solution framework ready for GenHMM, there are still remaining problems to address before it can be employed for practical usage, including
\begin{itemize}
\item how to realize GenHMM by neural network based generators such that likelihoond of their induced distributions can be computed explicitly and exactly?
\item how to train GenHMM to solve problem in \eqref{eq:ml-of-hmm} using practical algorithm?
\item would the training of GenHMM converge?
\end{itemize}
We tackle these problems in the following section.

@@

\subsection{Realizing $\bm{g}_{s,\kappa}$ by a Flow Model}
Each generator $\bm{g}_{s,\kappa}$ is realized as a feed-forward neural netowrk.
We define $\bm{g}_{s,\kappa}$ as a $L$-layer neural network and formulate its mapping by layer-wise concatenation:
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$, where superscript $[l]$ denotes the layer index and $\circ$ denotes mapping concatenation. Assume $\bm{g}_{s,\kappa}$ is invertible and denote its inverse mapping as $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$. For a latent variable $\bm{z}$ with density function $p_{s,\kappa}(\bm{z})$, the generated signal $\bm{x}$ follows an induced distribution with density function \eqref{eq:changel-variable}. We illustrate the signal flow between latent variable $\bm{z}$ and observable variable $\bm{x}$ as 
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{h}_L=\bm{x}}$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture}
\end{equation*}
where $\bm{f}_{s,\kappa}^{[l]}$ is the $l$-th layer of $\bm{f}_{s,\kappa}$. We have $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. If every layer of $\bm{g}_{s,\kappa}$ is invertible, the full feed-forward neural network is invertible. Flow model, proposed in \cite{DBLP:journals/corr/DinhKB14} as an image generating model, is such an invertible feed-forward layer-wise neural network. It is further improved in subsquential works \cite{2016arXiv160508803D,kingma2018glow} for high-fidelity and high-resolution image generating and representation. As shown in \eqref{eq:changel-variable}, the challenge lies at the computation of Jacobin determinant. {Another track of flow models uses a continuous-depth models instead. The variable change is defined by an ordinary differential equation implemented by a neural network \cite{NIPS2018_7892, DBLP:grathwohl2018FFJORD}, where the key becomes to solve the ODE problem. We use the layer-wise flow model to model the variable change in \eqref{eq:changel-variable} in which the efficient Jacobian computation is available.}

@@

Since $\bm{A}_{i, j}  = p(\smtr{s}{t+1}{r}=j|\smtr{s}{t}{r}=i; \bm{H})$ is the element of transition matrix $\bm{A}$, the solution to the problem
\begin{align}\label{eq:update-transition-prob}
  \bm{A} = &\uargmax{\bm{A}} \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) \nonumber \\
  \mathrm{s.t.} &\hspace{0.2cm} \bm{A} \cdot \bm{1} = \bm{1}, \bm{A}_{i,j} \geq 0 \,\, \forall i,j,
\end{align}
is
\begin{equation}
  \bm{A}_{i,j} = \frac{\bar{\xi}_{i,j}}{\sum_{k = 1}^{|\Ss|} \bar{\xi}_{i,k}},
\end{equation}
where
\begin{equation}\label{eq:update-transition-solt}
  \bar{\xi}_{i,j} = \sum_{r= 1}^{R} \sum_{t= 1}^{T^{r}-1}{p(\smtr{s}{t}{r}=i, \smtr{s}{t+1}{r}=j| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}.
\end{equation}
\subsection{On Convergence of GenHMM}
In pursuit of representing a dataset by GenHMM,  we are interested if the learning solution discussed in subsection~\ref{subsec:optmGenHMM} would convergence. The properties on GenHMM's convergence are analyzed as follows.

@@

\section{Experiments}
To show the validity of our model, we implement our model in PyTorch and test it with sequential data. We first discuss the experimental setups and then show the experimental results.

\subsection{Experimental Setups}
The dataset used for sequential data modeling and classification is TIMIT where the speech signal is sampled at $16$kHz.
The TIMIT dataset consists of $5300$ phoneme-labeled speech utterances which are partitioned into two sets: {a train set consists of $4620$ utterance, and a test set consists of $1680$ utterances.} There are totally $61$ different types of phones in TIMIT.
We performed experiments in two cases: i) full $61$-phoneme classification case; ii) $39$-phonme classification case, where $61$ phonemes are folded onto $39$ phonemes as described in \cite{Perdigao11}.

@@

\begin{table}
  \caption{Configuration of generators in Experiments}\label{table:generator-setting}
  \centering
  \begin{tabular}{l|c}
    \toprule
    \makecell{Latent distribution $p_{s,\kappa}(\bm{z})$
    \\$s\in \Ss, \kappa=1, 2, \cdots, K $} & Standard Gaussian \\
    \hline
    Number of flow blocks & $4$ \\
    \hline
    Non-linear mapping $\bm{m}_a$, $\bm{m}_b$ & \makecell{Multiple
                                                layer perception, \\
    $3$ layers and with hidden dimension $24$}\\
    \bottomrule
  \end{tabular}
\end{table}

@@

@@

\begin{table}
  \caption{Test accuracy table for $39$ dimensional features and $61$ phonemes.}\label{tab:acc-classification39f_b}
  \centering  
  \begin{tabular}{l|l|c|c|c} \toprule
    {Model} & Criterion & K=1 &  K=3 &  K=5
    \\ \midrule
    \multirow{4}{*}{GMM-HMM}
            & Accuracy    & $53.6\%$ &  $59.6\%$ & $61.9\%$  \\
            & {Precision} & $59.1\%$ &  $63.9\%$ & $65.7\%$ \\
            & {F1}        & $54.7\%$ &  $60.5\%$ & $62.7\%$\\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy    & $69.5\%$ & $70.6\%$ & $70.7\%$   \\
            & {Precision} & $69.2\%$ & $70.5\%$ & $71.0\%$ \\
            & {F1}        & $68.6\%$ & $69.6\%$ & $69.6\%$\\
    \bottomrule
  \end{tabular}
  \vspace{0.5cm}
\end{table}
We firstly show the phoneme classification using 39 dimensional MFCC features (MFCC coefficients, deltas, and delta-deltas), to validate one possible usage of our proposed model. Since generative training is carried out in our experiments, GMM-HMM is trained and tested as a reference model in our experiments. Training and testing of GMM-HMM is in the same condition as GenHMMs are trained and tested. Dataset usage for GenMM and GMM-HMM is the same, and number of states for GMM-HMM is the same as that for GenHMM in modeling each phoneme. Apart from setting the reference model, we also run the experiment comparisons with different total number of mixture components.

>>>>

With such a solution framework ready for GenHMM, there are still remaining problems to address before it can be employed for practical usage, including
\begin{itemize}
\item how to realize GenHMM by neural network based generators such that likelihood of their induced distributions can be computed explicitly and exactly?
\item how to train GenHMM to solve problem in \eqref{eq:ml-of-hmm} using practical algorithm?
\item would the training of GenHMM converge?
\end{itemize}
We tackle these problems in the following section.

@@

\subsection{Realizing $\bm{g}_{s,\kappa}$ by a Flow Model}
Each generator $\bm{g}_{s,\kappa}$ is realized as a feed-forward neural netowrk.
We define $\bm{g}_{s,\kappa}$ as a $L$-layer neural network and formulate its mapping by layer-wise concatenation:
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$, where superscript $[l]$ denotes the layer index and $\circ$ denotes mapping concatenation. Assume $\bm{g}_{s,\kappa}$ is invertible and denote its inverse mapping as $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$. For a latent variable $\bm{z}$ with density function $p_{s,\kappa}(\bm{z})$, the generated signal $\bm{x}$ follows an induced distribution with density function \eqref{eq:changel-variable}. We illustrate the signal flow between latent variable $\bm{z}$ and observable variable $\bm{x}$ as 
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{h}_L=\bm{x}}$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture}
\end{equation*}
where $\bm{f}_{s,\kappa}^{[l]}$ is the $l$-th layer of $\bm{f}_{s,\kappa}$. We have $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. If every layer of $\bm{g}_{s,\kappa}$ is invertible, the full feed-forward neural network is invertible. Flow model, proposed in \cite{DBLP:journals/corr/DinhKB14} as an image generating model, is such an invertible feed-forward layer-wise neural network. It is further improved in subsquential works \cite{2016arXiv160508803D,kingma2018glow} for high-fidelity and high-resolution image generating and representation. As shown in \eqref{eq:changel-variable}, the challenge lies at the computation of Jacobian determinant. {Another track of flow models uses a continuous-depth models instead. The variable change is defined by an ordinary differential equation implemented by a neural network \cite{NIPS2018_7892, DBLP:grathwohl2018FFJORD}, where the key becomes to solve the ODE problem. We use the layer-wise flow model to model the variable change in \eqref{eq:changel-variable} in which the efficient Jacobian computation is available.}

@@

Since $\bm{A}_{i, j}  = p(\smtr{s}{t+1}{r}=j|\smtr{s}{t}{r}=i; \bm{H})$ is the element of transition matrix $\bm{A}$, the solution to the problem
\begin{align}\label{eq:update-transition-prob}
  \bm{A} = &\uargmax{\bm{A}} \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) \nonumber \\
  \mathrm{s.t.} &\hspace{0.2cm} \bm{A} \cdot \bm{1} = \bm{1}, \bm{A}_{i,j} \geq 0 \,\, \forall i,j,
\end{align}
is
\begin{equation}
  \bm{A}_{i,j} = \frac{\bar{\xi}_{i,j}}{\sum_{k = 1}^{|\Ss|} \bar{\xi}_{i,k}},
\end{equation}
where
\begin{equation}\label{eq:update-transition-solt}
  \bar{\xi}_{i,j} = \sum_{r= 1}^{R} \sum_{t= 1}^{T^{r}-1}{p(\smtr{s}{t}{r}=i, \smtr{s}{t+1}{r}=j| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}.
\end{equation}
\subsection{On Convergence of GenHMM}
In pursuit of representing a dataset by GenHMM,  we are interested if the learning solution discussed in subsection~\ref{subsec:optmGenHMM} would converge. The properties on GenHMM's convergence are analyzed as follows.

@@

\section{Experiments}
To show the validity of our model, we implement our model in PyTorch and test it with sequential data. We first discuss the experimental setups and then show the experimental results. Code for experiments is available at {https://github.com/FirstHandScientist/genhmm}.

\subsection{Experimental Setup}
The dataset used for sequential data modeling and classification is TIMIT where the speech signal is sampled at $16$kHz.
The TIMIT dataset consists of $5300$ phoneme-labeled speech utterances which are partitioned into two sets: {a train set consists of $4620$ utterance, and a test set consists of $1680$ utterances.} There are totally $61$ different types of phones in TIMIT.
We performed experiments in two cases: i) full $61$-phoneme classification case; ii) $39$-phonme classification case, where $61$ phonemes are folded onto $39$ phonemes as described in \cite{Perdigao11}.

@@

\begin{table}
  \caption{Configuration of generators of GenHMM in Experiments}\label{table:generator-setting}
  \centering
  \begin{tabular}{l|c}
    \toprule
    \makecell{Latent distribution $p_{s,\kappa}(\bm{z})$
    \\$s\in \Ss, \kappa=1, 2, \cdots, K $} & Standard Gaussian \\
    \hline
    Number of flow blocks & $4$ \\
    \hline
    Non-linear mapping $\bm{m}_a$, $\bm{m}_b$ & \makecell{Multiple
                                                layer perception, \\
    $3$ layers and with hidden dimension $24$}\\
    \bottomrule
  \end{tabular}
\end{table}

@@

\begin{table}
  \caption{Test accuracy table for $39$ dimensional features and $61$ phonemes.}\label{tab:acc-classification39f_b}
  \centering  
  \begin{tabular}{l|l|c|c|c} \toprule
    {Model} & Criterion & K=1 &  K=3 &  K=5
    \\ \midrule
    \multirow{4}{*}{GMM-HMM}
            & Accuracy    & $53.6\%$ &  $59.6\%$ & $61.9\%$  \\
            & {Precision} & $59.1\%$ &  $63.9\%$ & $65.7\%$ \\
            & {F1}        & $54.7\%$ &  $60.5\%$ & $62.7\%$\\
    \midrule
    \multirow{3}{*}{GenHMM}
            & Accuracy    & $69.5\%$ & $70.6\%$ & $70.7\%$   \\
            & {Precision} & $69.2\%$ & $70.5\%$ & $71.0\%$ \\
            & {F1}        & $68.6\%$ & $69.6\%$ & $69.6\%$\\
    \bottomrule
  \end{tabular}
  \vspace{0.5cm}
\end{table}

@@

We firstly show the phoneme classification using 39 dimensional MFCC features (MFCC coefficients, deltas, and delta-deltas), to validate one possible usage of our proposed model. Since generative training is carried out in our experiments, GMM-HMM is trained and tested as a reference model in our experiments. Training and testing of GMM-HMM is in the same condition as GenHMMs are trained and tested. Dataset usage for GenMM and GMM-HMM is the same, and number of states for GMM-HMM is the same as that for GenHMM in modeling each phoneme. Apart from setting the reference model, we also run the experiment comparisons with different total number of mixture components.
